---
title: "Overall simulated batch characteristics"
author: "Almut Lütge"
date: "15 April 2020"
params:
  sce:
  sce_name:
  meta:
  summary:
  sce_norm:
output:
  html_document:
    self_contained: no
    lib_dir: '../../docs/site_libs'
    code_folding: hide
    theme: journal
    highlight: tango
    number_sections: no
    toc: yes
    toc_depth: 3
    toc_float: true
    collapsed: no
    smooth_scroll: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Batch Characteristics in simulated single cell RNAseq data

```{r libraries, warning=FALSE}
suppressPackageStartupMessages({
  library(scater)
  library(CellMixS)
  library(scales)
  library(purrr)
  library(circlize)
  library(viridis)
  library(jcolors)
  library(tidyr)
  library(dplyr)
  library(ComplexHeatmap)
  library(gridExtra)
  library(scran)
  library(magrittr)
  library(cowplot)
  library(colorspace)
})

```


Datasets

```{r data}
sce_name <- unlist(c(strsplit(params$sce_name, ",")))

sce_vec <- c(paste0(params$sce, sce_name, "_sce.rds"))
names(sce_vec) <- sce_name

meta_vec <- c(paste0(params$meta, sce_name, ".rds"))
names(meta_vec) <- sce_name

sce_small_vec <- c(paste0(params$sce_norm, "/", sce_name, "/sim_", sce_name, "_1__1_sce.rds"))
names(sce_small_vec) <- sce_name
sum_vec <- c(paste0(params$summary, sce_name, ".rds"))
names(sum_vec) <- sce_name


cols <-c(c(jcolors('pal6'),jcolors('pal8'))[c(1,8,14,5,2:4,6,7,9:13,15:20)],jcolors('pal4'))
names(cols) <- c()
```


### General 
General summaries of the datasets as number of cells, genes, batches, batch origin
```{r general}
general_info <- lapply(sce_name, function(sce_nam){
  sce <- readRDS(sce_vec[sce_nam])
  meta <- readRDS(meta_vec[sce_nam])
  d <- data.frame("n_cells" = ncol(sce),
                  "n_genes" = nrow(sce),
                  "n_batches" = length(levels(as.factor(colData(sce)[,meta$batch]))),
                  "n_celltypes" = length(levels(as.factor(colData(sce)[,meta$celltype]))),
                  "linear" = mean(sce$cms.Xadj1),
                  "additive" = mean(sce$cms.Xadj2),
                  "interacting" = mean(sce$cms.Xadj3),
                  "reference" = mean(sce$cms))
}) %>% set_names(sce_name) %>% bind_rows() %>% set_rownames(sce_name)

general_info$batch_origin <- rep(NA, nrow(general_info))
general_info[c("pbmc2_media", "pbmc_roche", "csf_media"), "batch_origin"] <- "media"
general_info[c("csf_patient", "kang", "pbmc2_pat"), "batch_origin"] <- "patient"
general_info[c("cellbench", "hca"), "batch_origin"] <- "protocol"
general_info$dataset <- rownames(general_info)

general_info <- general_info %>% arrange(batch_origin)
rownames(general_info) <- general_info$dataset

```

Plot general infos
```{r plot general infos}
#colors
q_type <- qualitative_hcl(4, palette = "Dark 3")
col_type <- list("protocol" = q_type[1], 
                     "media" = q_type[3], "patient" = q_type[4])

col_type_vec <- unlist(col_type[general_info$batch_origin])

q_batch <- sequential_hcl(5, palette = "BluYl")
col_batch <- colorRamp2(c(15, 10, 5, 3, 2), q_batch)

# annotations
h_cells <- HeatmapAnnotation(n_cells = anno_barplot(general_info$n_cells, 
                                                    gp = gpar(fill = col_type_vec),
                                                    height = unit(2.2, "cm")))

h_genes <- HeatmapAnnotation(n_genes = anno_barplot(general_info$n_genes, 
                                                    gp = gpar(fill = col_type_vec),
                                                    height = unit(2.2, "cm")))

# heatmap
n_batches <- t(as.matrix(general_info[,c("n_batches", "n_celltypes")])) %>% 
  set_colnames(general_info$dataset)


h_batch <- Heatmap(n_batches, 
          cluster_rows = FALSE, 
          cluster_columns = FALSE,
          name = "celltype/batch",
          column_names_side = "bottom",
          column_names_rot = 45,
          column_names_centered = TRUE,
          column_names_gp = gpar(fontsize = 8),
          row_names_side = "left", 
          col = col_batch,
          rect_gp = gpar(col= "white"),
          show_heatmap_legend = FALSE)



#legends
lgd1 = Legend(labels = names(col_type), legend_gp = gpar(fill = unlist(col_type)), 
              title = "batch origin")
lgd2 = Legend(col_fun = col_batch, at = c(0, 2, 5, 10, 15), 
              title = "celltype/batch")
pd = packLegend(lgd1, lgd2)

h_list <-  h_cells %v% h_genes %v% h_batch 

draw(h_list, annotation_legend_list = pd)

```

### Batch size/strength

```{r batch size}
batch_size <- lapply(sce_name, function(sce_nam){
  sum <- readRDS(sum_vec[sce_nam])
  meta <- readRDS(meta_vec[sce_nam])
  d <- data.frame("mean_var_batch" = sum$mean_var_batch,
                  "mean_var_celltype" = sum$mean_var_celltype,
                  "mean_cms" = sum$mean_cms,
                  "mean_de_genes" = sum$mean_mean_n_de_genes)
}) %>% set_names(sce_name) %>% bind_rows() %>% set_rownames(sce_name)

batch_size$batch_origin <- general_info[rownames(batch_size), "batch_origin"]
batch_size$dataset <- rownames(batch_size)
batch_size <- batch_size %>% arrange(batch_origin)
```


Plot batch size parameter
```{r plot batch size, fig.width=6, fig.height=3.5}
#colors
q_cms <- sequential_hcl(5, palette = "YlOrRd")
col_cms = colorRamp2(c(0, 0.1, 0.2, 0.3, 0.5), q_cms)

q_size <- sequential_hcl(5, palette = "Reds")
col_size <- colorRamp2(c(0.2, 0.1, 0.03, 0.01, 0), q_size)

col_type <- c("protocol" = q_type[1], "media" = q_type[3], "patient" = q_type[4])

q_size2 <- sequential_hcl(5, palette = "Blues")
col_size2 <- colorRamp2(c(0.65, 0.5, 0.3, 0.2, 0.1), q_size2)

#annotation
ha_type = HeatmapAnnotation("batch_origin" = batch_size$batch_origin,
                            simple_anno_size = unit(1, "cm"),
                            col = list("batch_origin" = col_type), 
                            annotation_height = unit(2.2, "cm"),
                            annotation_name_side = "left")

#heatmap
#cms
mean_cms <- t(as.matrix(batch_size$mean_cms))
colnames(mean_cms) <- batch_size$dataset
rownames(mean_cms) <- "mean_cms"

h_cms <- Heatmap(mean_cms, 
          cluster_rows = FALSE, 
          cluster_columns = FALSE,
          name = "cms",
          column_names_side = "bottom",
          column_names_rot = 45,
          column_names_centered = TRUE,
          column_names_gp = gpar(fontsize = 8),
          row_names_side = "left", 
          col = col_cms,
          rect_gp = gpar(col= "white"))

# mean variance part
n_batches <- t(as.matrix(batch_size[,c("mean_var_batch", "mean_var_celltype")])) %>% 
  set_colnames(batch_size$dataset)

h_size1 <- Heatmap(n_batches, 
          cluster_rows = FALSE, 
          cluster_columns = FALSE,
          name = "variance",
          column_names_side = "bottom",
          column_names_rot = 0,
          column_names_centered = TRUE,
          column_names_gp = gpar(fontsize = 8),
          row_names_side = "left", 
          col = col_size,
          top_annotation = ha_type,
          rect_gp = gpar(col= "white"))


mean_de_genes <- t(as.matrix(batch_size$mean_de_genes))
colnames(mean_de_genes) <- batch_size$dataset
rownames(mean_de_genes) <- "mean_de_genes"

h_size2 <- Heatmap(mean_de_genes, 
          cluster_rows = FALSE, 
          cluster_columns = FALSE,
          name = "de genes [%]",
          column_names_side = "bottom",
          column_names_rot = 0,
          column_names_centered = TRUE,
          column_names_gp = gpar(fontsize = 8),
          row_names_side = "left", 
          col = col_size2,
          rect_gp = gpar(col= "white"))

h_list <-  h_size1 %v% h_size2 %v% h_cms

draw(h_list)

```



### Batch type

```{r batch type}
#cols
q_cms <- sequential_hcl(5, palette = "Purple-Blue")
col_cms = colorRamp2(c(0, 0.1, 0.2, 0.3, 0.5), q_cms)

#annotation
ha_type = HeatmapAnnotation("batch_origin" = batch_size$batch_origin, 
                            simple_anno_size =  unit(1, "cm"),
                            col = list("batch_origin" = col_type), 
                            annotation_height = unit(2.2, "cm"),
                            annotation_name_side = "left")

#heatmap
#cms
type <- t(as.matrix(general_info[,c("reference","linear", "additive", "interacting")])) %>% 
  set_colnames(batch_size$dataset)

h_cms <- Heatmap(type, 
          cluster_rows = FALSE, 
          cluster_columns = FALSE,
          name = "cms",
          column_names_side = "bottom",
          column_names_rot = 45,
          column_names_centered = TRUE,
          column_names_gp = gpar(fontsize = 8),
          row_names_side = "left", 
          col = col_cms,
          top_annotation = ha_type,
          rect_gp = gpar(col= "white"))

draw(h_cms)
```



### Celltype specificity

#### Variables

```{r ct spec}
ct_spec <- lapply(sce_name, function(sce_nam){
  sum <- readRDS(sum_vec[sce_nam])
  meta <- readRDS(meta_vec[sce_nam])
  d <- data.frame("celltype_var_cms" = sum$celltype_var_cms,
                  "mean_lfc_cor" = sum$mean_lfc_cor,
                  "mean_de_overlap" = sum$mean_de_overlap)
}) %>% set_names(sce_name) %>% bind_rows() %>% set_rownames(sce_name)

ct_spec$batch_origin <- general_info[rownames(ct_spec), "batch_origin"]
ct_spec$dataset <- rownames(ct_spec)
ct_spec <- ct_spec %>% arrange(batch_origin)
```


Plot celltype specificity
```{r plot ct spec}
#colors
q_cms <- sequential_hcl(5, palette = "BluYl")
col_cms = colorRamp2(c(0, 0.002, 0.004, 0.006, 0.008), q_cms)

q_cor <- rev(sequential_hcl(6, palette = "Reds"))
col_cor <- colorRamp2(c(0, 0.2, 0.4, 0.6, 0.8, 0.9), q_cor)

col_type <- c("protocol" = q_type[1], "media" = q_type[3], "patient" = q_type[4])

q_over <- rev(sequential_hcl(6, palette = "Blues"))
col_over <- colorRamp2(c(0, 0.1, 0.2, 0.3, 0.4, 0.5), q_over)

#annotation
ha_type = HeatmapAnnotation("batch_origin" = ct_spec$batch_origin,
                            simple_anno_size = unit(1, "cm"),
                            col = list("batch_origin" = col_type), 
                            annotation_height = unit(2.2, "cm"),
                            annotation_name_side = "left")

#heatmap
#cms
var_cms <- t(as.matrix(ct_spec$celltype_var_cms))
colnames(var_cms) <- ct_spec$dataset
rownames(var_cms) <- "var_cms"

h_cms <- Heatmap(var_cms, 
          cluster_rows = FALSE, 
          cluster_columns = FALSE,
          name = "cms variance",
          column_names_side = "bottom",
          column_names_rot = 45,
          column_names_centered = TRUE,
          column_names_gp = gpar(fontsize = 8),
          row_names_side = "left", 
          col = col_cms,
          rect_gp = gpar(col= "white"))

# correlation
cor_tab <- t(as.matrix(ct_spec$mean_lfc_cor))
colnames(cor_tab) <- ct_spec$dataset
rownames(cor_tab) <- "lfc_correlation"

h_cor <- Heatmap(cor_tab, 
          cluster_rows = FALSE, 
          cluster_columns = FALSE,
          name = "lfc correlation",
          column_names_side = "bottom",
          column_names_rot = 0,
          column_names_centered = TRUE,
          column_names_gp = gpar(fontsize = 8),
          row_names_side = "left", 
          col = col_cor,
          top_annotation = ha_type,
          rect_gp = gpar(col= "white"))


over <- t(as.matrix(ct_spec$mean_de_overlap))
colnames(over) <- ct_spec$dataset
rownames(over) <- "de_overlap"

h_over <- Heatmap(over, 
          cluster_rows = FALSE, 
          cluster_columns = FALSE,
          name = "de overlap [%]",
          column_names_side = "bottom",
          column_names_rot = 0,
          column_names_centered = TRUE,
          column_names_gp = gpar(fontsize = 8),
          row_names_side = "left", 
          col = col_over,
          rect_gp = gpar(col= "white"))

h_list <-  h_cor %v% h_over %v% h_cms

draw(h_list)
```


#### Distributions
```{r get lfc}
gene_info <- lapply(sce_name, function(sce_nam){
  gc()
  sce <- readRDS(sce_vec[sce_nam])
  meta <- readRDS(meta_vec[sce_nam])
  d <- rowData(sce)[, grep("logFC", names(rowData(sce)))]
}) %>% set_names(sce_name)


ct_info <- lapply(sce_name, function(sce_nam){
  sce <- readRDS(sce_small_vec[sce_nam])
  meta <- readRDS(meta_vec[sce_nam])
  ct <- table(colData(sce)[, meta$celltype])
}) %>% set_names(sce_name)

```

### logFC distributions {.tabset}
```{r plot lfc}

plot_den <- function(nam){
  mat <- as.matrix(gene_info[[nam]])
  ct <- ct_info[[nam]]
  #col and factors
  ba_tab <- data.frame("batch" = gsub("_.*","",colnames(mat)),
                     "celltype" = gsub(".*_logFC_", "", colnames(mat)))
  ba_tab$cells <- ct[ba_tab$celltype]
  col_den_b <- rev(cols)[3:(2 + length(levels(as.factor(ba_tab$batch))))] %>% 
  set_names(levels(as.factor(ba_tab$batch)))
  col_den_c <- cols[1:length(levels(as.factor(ba_tab$celltype)))] %>% 
  set_names(levels(as.factor(ba_tab$celltype)))
  q_cells <- sequential_hcl(3, palette = "Red-Blue")
  col_cells <- colorRamp2(c(max(ct), mean(ct), min(ct)), q_cells)
  #annotations
  ha_den = HeatmapAnnotation("batch" = ba_tab$batch,
                            "celltype" = ba_tab$celltype,
                            "cells" = ba_tab$cells,
                            simple_anno_size = unit(1, "cm"),
                            col = list("batch" = col_den_b,
                                       "celltype" = col_den_c,
                                       "cells" = col_cells),
                            annotation_name_side = "left")
  #heatmap
  m <- densityHeatmap(mat, 
                    ylim = c(-2, 2),
                    title = paste0("Density heatmap of logFC by celltype: ", nam),
                    top_annotation = ha_den,
                    show_column_names = FALSE
                    )
  draw(m)
}


template_density <- c(
    "#### {{nam}}\n",
    "```{r density{{nam}}, echo = FALSE}\n",
    "m <- plot_den('{{nam}}')",
    "```\n",
    "\n"
  )

plots_dens<- lapply(sce_name, 
  function(nam) knitr::knit_expand(text = template_density)
)

```

`r knitr::knit(text = unlist(plots_dens))`


