---
title: "Characterize simulated batch effects"
author: "Almut Lütge"
date: "`r format(Sys.time(), '%d %B, %Y')`"
params:
  data:
  param:
  de:
  gs:
  abund:
  out_file:
output:
  html_document:
    self_contained: no
    lib_dir: '../../docs/site_libs'
    code_folding: show
    theme: journal
    highlight: tango
    number_sections: no
    toc: yes
    toc_depth: 3
    toc_float: true
    collapsed: no
    smooth_scroll: yes
---


```{r dataset name, include=FALSE}
param <- readRDS(params$param)
dataset_name <- param[["dataset_name"]]

```

## `r paste(dataset_name)`

[Back to home](index.html)

```{r libraries, warning=FALSE}
suppressPackageStartupMessages({
  library(CellBench)
  library(scater)
  library(CellMixS)
  library(variancePartition)
  library(purrr)
  library(jcolors)
  library(here)
  library(tidyr)
  library(dplyr)
  library(gridExtra)
  library(stringr)
  library(ComplexHeatmap)
  library(scran)
  library(cowplot)
  library(CAMERA)
  library(ggrepel)
  library(readr)
})
```


### Dataset and parameter
```{r data}
sce <- readRDS(params$data)

param <- readRDS(params$param)
celltype <- param[["celltype"]]
batch <- param[["batch"]]
sample <- param[["Sample"]]
dataset_name <- param[["dataset_name"]]
dataset_name
n_genes <- nrow(sce)

table(colData(sce)[,celltype])
table(colData(sce)[,batch])
table(colData(sce)[,sample])

res_de <- readRDS(params$de)
abund <- readRDS(params$abund)
outputfile <- params$out_file

cols <-c(c(jcolors('pal6'),jcolors('pal8'))[c(1,8,14,5,2:4,6,7,9:13,15:20)],jcolors('pal4'))
names(cols) <- c()
```

### Visualize data
How are sample, celltypes and batches distributed within normalized, but not batch corrected data?
```{r vis}
feature_list <- c(batch, celltype, sample)
feature_list <- feature_list[which(!is.na(feature_list))]

lapply(feature_list, function(feature_name){
  visGroup(sce, feature_name, dim_red= "UMAP")
})
```

## Batch strength/size

To compare or describe the severity of a batch effect there are different meassures.
In general they can either give an estimate of the relative strength compared to the 
signal of interest e.g. the celltype signal or an absolut estimate e.g. the number 
of batch affected genes.    
  
### Variance partitioning 

How much of the variance within the datasets can we attributed to the batch effect and how much could be explained by the celltype? Which genes are mostly affected?

```{r variance part}
vp_vars <- c("vp_batch", "vp_celltype", "vp_residuals")
vp <- as_tibble(rowData(sce)[, vp_vars])  %>% dplyr::mutate(gene= rownames(sce)) %>% dplyr::arrange(-vp_batch)
vp_sub <- vp[1:3] %>% set_rownames(vp$gene)

#plot
plotPercentBars( vp_sub[1:10,] )
plotVarPart( vp_sub )
```

### Variance and gene expression {.tabset .tabset-pills}

Are general expression and batch effect related? Does the batch effect or the celltype effect preferable manifest within highly, medium or low expressed genes?

```{r var expr}

#define expression classes by mean expression quantiles 
th <- quantile(rowMeans(assays(sce)$logcounts), c(.33, .66))
high_th <- th[2]
mid_th <- th[1]

rowData(sce)$expr_class <- ifelse(rowMeans(assays(sce)$logcounts) > high_th, "high",
                                  ifelse(rowMeans(assays(sce)$logcounts) <= high_th &
                                           rowMeans(assays(sce)$logcounts) > mid_th, 
                                         "medium", "low"))
rowData(sce)$mean_expr <- rowMeans(assays(sce)$logcounts)

#plot 
plot_dev <- function(var, var_col){
  ggplot(as.data.frame(rowData(sce)), aes_string(x = "mean_expr", y = var, colour = var_col)) + 
  geom_point() + 
  geom_smooth(method = "lm", se = FALSE)
}

#Ternary plots
# ggtern(data=as.data.frame(rowData(sce)),aes(vp_batch, vp_celltype, vp_residuals)) + 
#   stat_density_tern(aes(fill=..level.., alpha=..level..),geom='polygon') +
#   scale_fill_gradient2(high = "red") +
#   guides(color = "none", fill = "none", alpha = "none") +
#   geom_point(size= 0.1, alpha = 0.5)  + 
#   Llab("batch") +
#   Tlab("celltype") +
#   Rlab("other") +
#   theme_bw()
# 
# t1 <- ggtern(data=as.data.frame(rowData(sce)),aes(vp_batch, vp_celltype, vp_residuals)) + 
#   geom_point(size = 0.1) +
#   geom_density_tern() + 
#   Llab("batch") +
#   Tlab("celltype") +
#   Rlab("other") +
#   theme_bw()

## Summarize variance partitioning
# How many genes have a variance component affected by batch with > 1%
n_batch_gene <- vp_sub %>% dplyr::filter(vp_batch > 0.01) %>% nrow()/n_genes
n_batch_gene10 <- vp_sub %>% dplyr::filter(vp_batch > 0.1) %>% nrow()/n_genes
n_celltype_gene <- vp_sub %>% dplyr::filter(vp_celltype> 0.01) %>% nrow()/n_genes
n_rel <- n_batch_gene/n_celltype_gene

# Mean variance that is explained by the batch effect/celltype
m_batch <- mean(vp_sub$vp_batch, na.rm = TRUE)
m_celltype <- mean(vp_sub$vp_celltype, na.rm = TRUE)
m_rel <- m_batch/m_celltype

```

#### Scatterplot batch
```{r var_plot_batch, warning=FALSE}
plot_dev("vp_batch", "vp_batch")
```

#### Scatterplot celltype
```{r var_plot_celltype, warning=FALSE}
plot_dev("vp_celltype", "vp_celltype")
```


#### Ternary plot all genes
```{r tern plots, warning=FALSE}
#t1

```

#### Ternary plot gene expression classes
```{r tern plots gene class, fig.asp=0.5, warning=FALSE}
#t1 + facet_grid(~expr_class)
```


### Cellspecific Mixing score  {.tabset .tabset-pills}

#### Overall
```{r cms}
#visualize overall cms score
visHist(sce, n_col = 2, prefix = FALSE)
visMetric(sce, metric = "cms_smooth", dim_red = "UMAP")
visGroup(sce, celltype, dim_red = "UMAP")

#summarize
mean_cms <- mean(sce$cms)
n_cms_0.01 <- length(which(sce$cms < 0.01))
cluster_mean_cms <- as_tibble(colData(sce)) %>% group_by_at(celltype) %>% summarize(cms_mean = mean(cms))
var_cms <- var(cluster_mean_cms$cms_mean)

```

#### Celltypes cms smooth
```{r cms celltypes}
#compare by celltypes
visCluster(sce, metric_var = "cms_smooth", cluster_var = celltype)
visCluster(sce, metric_var = "cms_smooth", cluster_var = celltype, violin = TRUE)
```

#### Celltypes histogram
```{r cms histogram}
#compare histogram by celltype

p <- ggplot(as.data.frame(colData(sce)), 
            aes_string(x = "cms", fill = celltype)) + 
                geom_histogram() + 
                facet_wrap(celltype, scales = "free_y", ncol = 3) +
                scale_fill_manual(values = cols) +
                theme_classic()

p + geom_vline(aes_string(xintercept = "cms_mean", 
                   colour = celltype), 
               cluster_mean_cms, linetype=2) + 
  scale_color_manual(values = cols) 
```


## Celltype specificity

### Celltype abundance
```{r celltype abundance}
meta_tib <- as_tibble(colData(sce)) %>% group_by_at(c(batch, celltype)) %>% summarize(n = n()) %>% dplyr::mutate(cell_freq = n / sum(n))

plot_abundance <- function(cluster_var, tib, x_var){
  meta_df <- as.data.frame(eval(tib))
  p <- ggplot(data=meta_df, aes_string(x=x_var, y="cell_freq", fill = cluster_var)) +
    geom_bar(stat="identity") + scale_fill_manual(values=cols, name = "celltype")
  p + coord_flip() + theme_minimal() 
}

plot_abundance(cluster_var = celltype, tib = meta_tib, x_var = batch)

#summarize diff abundance
mean_rel_abund_diff <- mean(unlist(abund))
min_rel_abund_diff <- min(unlist(abund))
max_rel_abund_diff <- max(unlist(abund))

```

### Batch and celltype specific count distributions
Do the overall count distribution vary between batches? Are count distributions celltype depended
  
```{r count distributions, fig.width = 10, fig.height = 9}
#batch level
bids <- levels(as.factor(colData(sce)[, batch]))
names(bids) <- bids
cids <- levels(as.factor(colData(sce)[, celltype]))
names(cids) <- cids

#mean gene expression by batch and cluster
mean_list <- lapply(bids, function(batch_var){
  mean_cluster <- lapply(cids, function(cluster_var){
    counts_sc <- as.matrix(logcounts(
      sce[, colData(sce)[, batch] %in% batch_var & 
            colData(sce)[, celltype] %in% cluster_var]))
  })
  mean_c <- mean_cluster %>% map(rowMeans) %>% bind_rows %>%
    dplyr::mutate(gene=rownames(sce)) %>% 
    gather(cluster, logcounts, cids)
})

mean_expr <- mean_list %>% bind_rows(.id= "batch")

ggplot(mean_expr, aes(x=logcounts, colour=batch)) + geom_density(alpha=.3) +
  theme_classic() +
  facet_wrap( ~ cluster, ncol = 3) +
  scale_colour_manual(values = cols[c(1:3,7,4:6,8:length(cols))]) +
  scale_x_continuous(limits =  c(0, 7))

```

### Batch to batch comparisons of expression distributions
```{r mean expression, eval=FALSE, fig.height=8, include=FALSE}

#mean expression
mean_expr <- mean_list %>% bind_rows(.id = "batch" ) %>% spread(batch, logcounts)
batch_all <- levels(as.factor(colData(sce)[,batch]))

lapply(batch_all, function(batch_var){
  batch_var_2 <- batch_all[-which(batch_all %in% batch_var)]
  lapply(batch_var_2, function(batch_var_3){
    ggplot(mean_expr, aes_string(x=batch_var, y = batch_var_3)) +
      geom_point(alpha = .3, aes(color=cluster)) +
      ggtitle(batch_var) + 
      geom_abline(slope = 1) + 
      coord_fixed() +
      facet_wrap( ~ cluster, ncol = 3) +
      scale_color_manual(values=cols) +
      theme_classic()
  })
})

```

## Differentially expressed genes
### Upset plot
```{r de res}
## Upset plot\
cont <- param[["cont"]]
cs <- names(cont)
names(cs) <- cs

# Filter DEG by pvalue
FilterDEGs <- function (degDF = df, filter = c(FDR = 5)){
  rownames(degDF) <- degDF$gene
  #pval <- degDF[, grep("adj.P.Val$", colnames(degDF)), drop = FALSE]
  pval <- degDF[, grep("PValue$", colnames(degDF)), drop = FALSE]
  pf <- pval <= filter["FDR"]/100
  pf[is.na(pf)] <- FALSE
  DEGlistUPorDOWN <- sapply(colnames(pf), function(x) rownames(pf[pf[, x, drop = FALSE], , drop = FALSE]), simplify = FALSE)
}

result <- list()
m2 <- list()

for(jj in 1:length(cs)){
  result[[jj]] <- sapply(res_de[[1]][[names(cs)[jj]]], function(x) FilterDEGs(x))
  names(result[[jj]]) <- cids
  m2[[jj]] = make_comb_mat(result[[jj]], mode = "intersect")
}

names(result) <- names(cs)
names(m2) <- names(cs)

lapply(m2, function(x) UpSet(x))

```


### Logfold_change 
```{r pval lfc}
# DE genes (per cluster and mean)
res <- res_de[["table"]]
#n_de <- lapply(res, function(y) vapply(y, function(x) sum(x$adj.P.Val < 0.05), numeric(1)))
n_de <- lapply(res, function(y) vapply(y, function(x) sum(x$PValue < 0.05), numeric(1)))
n_genes_lfc1 <- lapply(res, function(y) vapply(y, function(x) sum(abs(x$logFC) > 1), numeric(1)))
mean_n_genes_lfc1 <- mean(unlist(n_genes_lfc1))/n_genes

# plot DE for all comparison 
gs <- read_delim(params$gs, delim = "\n", col_names = "cat")
cats <- sapply(gs$cat, function(u) strsplit(u, "\t")[[1]][-2], 
               USE.NAMES = FALSE)
names(cats) <- sapply(cats, .subset, 1)
cats <- lapply(cats, function(u) u[-1])

plotDE <- function(cont_var){
  #res_s <- res[[cont_var]] %>% map(filter, adj.P.Val < .05) %>% map(filter, abs(logFC) > 1)
  res_s <- res[[cont_var]] %>% map(filter, PValue < .05) %>% map(filter, abs(logFC) > 2)
  
  #plot
  lapply(names(res[[cont_var]]), function(ct){
    ct_de <- res[[cont_var]][[ct]]
    ct_de$gene <- gsub('[A-z0-9]*\\.', '', ct_de$gene)
    res_s[[ct]]$gene <- gsub('[A-z0-9]*\\.', '', res_s[[ct]]$gene)
    
    #p <- ggplot(ct_de, aes(x = AveExpr, y = logFC, colour = abs(logFC) > 1, label = gene)) + 
    p <- ggplot(ct_de, aes(x = logCPM, y = logFC, colour = abs(logFC) > 1, label = gene)) +
      geom_point(size = 2, alpha = .5) +
      geom_label_repel(data = res_s[[ct]]) +
      ggtitle(paste0(ct,": ", cont_var)) +
      theme_classic()
    print(p)
    
    cat("Cluster:", ct, "Contrast:", cont_var, 
        "Num genes:", nrow(ct_de), "Num DE:", nrow(res_s[[ct]]), "\n" )

  })
}

if( length(names(res)) <= 3 ){
    pathways <- lapply(names(res), plotDE)
}

```



Summarize differential expression analysis
```{r summarize de}
# DE genes (per cluster and mean)
#n_de <- lapply(res, function(y) vapply(y, function(x) sum(x$adj.P.Val < 0.05), numeric(1)))
n_de <- lapply(res, function(y) vapply(y, function(x) sum(x$PValue < 0.05), numeric(1)))
n_de_cl <- lapply(res, function(y) vapply(y, function(x) nrow(x), numeric(1)))
mean_n_de <- lapply(n_de, function(x) mean(x))
mean_mean_n_de <- mean(unlist(mean_n_de))/n_genes
min_mean_n_de <- min(unlist(mean_n_de))/n_genes
max_mean_n_de <- max(unlist(mean_n_de))/n_genes

# Genes with lfc > 1
n_genes_lfc1 <- lapply(res, function(y) vapply(y, function(x) sum(abs(x$logFC) > 1), numeric(1)))
mean_n_genes_lfc1 <- mean(unlist(n_genes_lfc1))/n_genes
min_n_genes_lfc1 <- min(unlist(n_genes_lfc1))/n_genes
max_n_genes_lfc1 <- max(unlist(n_genes_lfc1))/n_genes

# DE genes overlap between celltypes (celltype specific de genes)
# Genes are "overlapping" if they are present in all clusters with at least 10% of all cells
de_overlap <- lapply(result, function(x){
  result2 <- x[table(colData(sce)[, celltype]) > ncol(sce) * 0.1]
  de_overlap <- length(Reduce(intersect, result2))
  de_overlap
})

mean_de_overlap <- mean(unlist(de_overlap))/n_genes
min_de_overlap <- min(unlist(de_overlap))/n_genes
max_de_overlap <- max(unlist(de_overlap))/n_genes

#Genes unique to single celltypes
unique_genes_matrix <- NULL
unique_genes <- NULL
cb <- length(names(result[[1]]))
unique_genes <- lapply(result,function(x){
  for( i in 1:cb ){
    unique_genes[i] <-as.numeric(length(setdiff(unlist(x[i]),unlist(x[-i]))))
  }
  unique_genes_matrix <- cbind(unique_genes_matrix, unique_genes)
  unique_genes_matrix
})

unique_genes <- Reduce('cbind', unique_genes)
colnames(unique_genes) <- names(result)
rownames(unique_genes) <- names(result[[1]])

# Relative cluster specificity (unique/overlapping)
rel_spec1 <- NULL
for( i in 1:dim(unique_genes)[2] ){
  rel_spec <- unique_genes[,i]/de_overlap[[i]]
  rel_spec1 <- cbind(rel_spec1,rel_spec)
}

mean_rel_spec <- mean(rel_spec1)
min_rel_spec <- min(rel_spec1)
max_rel_spec <- max(rel_spec1)

```

### Celltype specific DE distributions

How similar is the batch effect between celltypes. 
Do we have similar logFC distributions or different?
```{r ct de dist , fig.width=20, fig.height=20}

combine_folds <- function(cont_var){
  #extract the contrast of interest and change log2fold colums names to be unique
  B <- res[[cont_var]] 
  new_name <- function(p){
    colnames(B[[p]])[3] <- paste0("logFC_", p)
    return(B[[p]][,c(1,3)])
    }
  B_new_names <- lapply(names(B),new_name)
  names(B_new_names) <- names(B)
  #combine log2fold colums
  Folds <- Reduce(function(...){inner_join(..., by="gene")}, B_new_names)
}

all_folds <- lapply(cs, combine_folds)

#define pannels for pairs() function
panel.cor <- function(x, y, digits = 2, cex.cor){
  usr <- par("usr"); on.exit(par(usr))
  par(usr = c(0, 1, 0, 1))
  r <- abs(cor(x, y))
  txt <- format(c(r, 0.123456789), digits=digits)[1]
  test <- cor.test(x,y)
  Signif <- ifelse(round(test$p.value, 3) < 0.001,
                   "p<0.001",
                   paste("p=",round(test$p.value,3)))  
  text(0.5, 0.25, paste("r=",txt), cex = 3)
  text(.5, .75, Signif, cex = 3)
}

panel.smooth <- function (x, y, col = "blue", bg = NA, pch = 18, cex = 1.5, 
                          col.smooth = "red", span = 2/3, iter = 3, ...){
  points(x, y, pch = pch, col = col, bg = bg, cex = cex)
  ok <- is.finite(x) & is.finite(y)
  if( any(ok) ) 
    lines(stats::lowess(x[ok], y[ok], f = span, iter = iter), 
          col = col.smooth, ...)
}

panel.hist <- function(x, ...){
  usr <- par("usr"); on.exit(par(usr))
  par(usr = c(usr[1:2], 0, 1.5) )
  h <- hist(x, plot = FALSE)
  breaks <- h$breaks
  nB <- length(breaks)
  y <- h$counts
  y <- y/max(y)
  rect(breaks[-nB], 0, breaks[-1], y, col="cyan", ...)
}
#plot correlations
lapply(names(all_folds), function(x) pairs(all_folds[[x]][,-1],
                                           lower.panel = panel.smooth, 
                                           upper.panel = panel.cor, 
                                           diag.panel = panel.hist, main = x))

#extract correlation coefficients
# correlation coefficients from celltype specific gege logFC

lfc_cor_list <-lapply(names(all_folds), function(com){
  exclude <- which(table(colData(sce)[,celltype]) < 100)
  r <- cor(all_folds[[com]][, -c(1, (exclude + 1))])
  mean_r <- (sum(r) - ncol(r))/ (ncol(r)^2 - ncol(r))
})

mean_lfc_cor <- mean(unlist(lfc_cor_list))

```


Compare estimated logFC between real data and simulated data.
This can give as a feeling for how much we underestimate the batch effect. 
In extremer case the batch tuning parameters can be used to get a batch effect 
closer to the real batch effect.

```{r compare lfcs, warning=FALSE, message=FALSE}
comp_lfcs <- function(cont_var){
    print(cont_var)
    lapply(cids, function(cell_t){
        print(cell_t)
        c1 <- gsub("-.*", "", cont_var)
        c2 <- gsub(".*-", "", cont_var)
        cont_fc <- colnames(sce@metadata[["gene_info"]]) %>% 
            grep(paste0("lfc_be_", c2), ., value = TRUE)
        sim_est <- all_folds[[cont_var]][, c("gene", paste0("logFC_", cell_t))]
        real_est <- as_tibble(sce@metadata[["gene_info"]]) %>% 
            filter(cluster_id %in% cell_t) %>% select(c("gene", all_of(cont_fc)))
        if( ncol(real_est) < 2 ){
            p <- NULL
        }else{
            lfc_com <- full_join(real_est, sim_est) %>% set_colnames(c("gene", "real", "simulated"))
        p <- ggplot(lfc_com, aes(y=real, x=simulated)) + 
            geom_point(alpha = 0.3, color='darkblue') +
            geom_abline(intercept = 0, slope = 1) +
            labs(title=paste0("Compare logFC estimates: ", cont_var, ", cluster: ", cell_t), 
                 x = "logFC simulated", y = "logFC real") + 
            theme_classic() + coord_fixed()
        p
        }
    })
}

lapply(cs, comp_lfcs)
```


## Batch categorization {.tabset .tabset-pills}

How does the batch effect manifest? Can we describe it by “simple” mean shifts of expression levels for some genes for all the cells in a given celltype and batch? Can we "remove" the batch effcet using 
a linear model with batch, batch and celltype or batch and celltype interacting?

```{r batch types}
#Visualize different models
vis_type <- function(dim_red){
  g <- visGroup(sce, batch, dim_red = dim_red) + 
    ggtitle("unadjusted")
  g1 <- visGroup(sce, batch, dim_red = paste0(dim_red, "_Xadj1")) + 
    ggtitle("constant batch effect")
  g2 <- visGroup(sce, batch, dim_red = paste0(dim_red, "_Xadj2")) + 
    ggtitle("constant batch effect, different ct composition")
  g3 <- visGroup(sce, batch, dim_red = paste0(dim_red, "_Xadj3")) + 
    ggtitle("celltype and batch effect interact")
  do.call("grid.arrange", c(list(g, g1, g2, g3), ncol = 2))
}



```

### PCA
```{r batchtype pca}
vis_type("PCA")
```

### UMAP
```{r batchtype umap}
vis_type("UMAP")
```

### Cellspecific Mixing Score
```{r batchtype cms}
# #Cellspecific Mixing score (Batch effect strength after "removal")
visHist(sce, metric = c("cms", "cms.Xadj1", "cms.Xadj2", "cms.Xadj3"), prefix = FALSE)
visIntegration(sce, metric = c("cms", "cms.Xadj1", "cms.Xadj2", "cms.Xadj3"), metric_name = "cms", prefix = FALSE)
```


## Simulation parameter
Extract parameter to use as input into simualation
```{r sim_param}
#percentage of batch affected genes
cond <- gsub("-.*", "", names(n_de))
cond <- c(cond, unique(gsub(".*-", "", names(n_de))))
cond <- unique(cond)
de_be_tab <- n_de %>% bind_cols()
de_cl_tab <- n_de_cl %>% bind_cols()

de_be <- cond %>% map(function(x){
  de_tab <- de_be_tab[, grep(x, colnames(de_be_tab))]
  de_be <- rowMeans(de_tab)
}) %>% bind_cols() %>% set_colnames(cond)

n_cl <- cond %>% map(function(x){
  cl_tab <- de_cl_tab[, grep(x, colnames(de_cl_tab))]
  de_cl <- rowMeans(cl_tab)
}) %>% bind_cols() %>% set_colnames(cond)


p_be <- de_be/n_cl
mean_p_be <- mean(colMeans(p_be))
min_p_be <- min(colMins(as.matrix(p_be)))
max_p_be <- max(colMaxs(as.matrix(p_be)))
sd_p_be <- mean(colSds(as.matrix(p_be)))
if(is.na(sd_p_be)){ sd_p_be <- 0 }


#### Percentage of celltype specific genes "p_ct"
n_de_unique <- lapply(result,function(x){
  de_genes <- unlist(x) %>% unique() %>% length()
  de_genes <- de_genes/length(x)
}) %>% bind_cols()


rel_spec2 <- NULL
for(i in 1:length(de_overlap)){
  rel_spec <- de_overlap[[i]]/mean(n_de[[i]][table(colData(sce)[, celltype]) > dim(expr)[2] * 0.1])
  rel_spec2 <- cbind(rel_spec2, rel_spec)
}

mean_p_ct <- 1 - mean(rel_spec2)
max_p_ct <- 1 - min(rel_spec2)
min_p_ct <- 1 - max(rel_spec2)
sd_p_ct <- sd(rel_spec2)
if(is.na(sd_p_ct)){ sd_p_ct <- 0 }

# Logfold change
#logFoldchange batch effect distribution
mean_lfc_cl <- lapply(res, function(y) vapply(y, function(x){
  #de_genes <- which(x$adj.P.Val < 0.05)
  de_genes <- which(x$PValue < 0.05)
  mean_de <- mean(abs(x[, "logFC"]))}
  , numeric(1))) %>% bind_cols()

mean_lfc_be <- mean(colMeans(mean_lfc_cl, na.rm = TRUE))
min_lfc_be <- min(colMins(as.matrix(mean_lfc_cl), na.rm = TRUE))
max_lfc_be <- max(colMaxs(as.matrix(mean_lfc_cl), na.rm = TRUE))

```


## Summarize batch effect
- Batch size
- Celltype specificity
- "Batch genes"
- batch type

```{r summarize }
#Size? How much of the variance can be attributed to the batch effect?
size <- data.frame("batch_genes_1per" = n_batch_gene,       # 1.variance partition
                   "batch_genes_10per" = n_batch_gene10,
                   "celltype_gene_1per" = n_celltype_gene,
                   "relative_batch_celltype" = n_rel,
                   "mean_var_batch" = m_batch,
                   "mean_var_celltype" = m_celltype,
                   "rel_mean_ct_batch" = m_rel,
                   "mean_cms" = mean_cms,                    #2.cms
                   "n_cells_cms_0.01" = n_cms_0.01,
                   "mean_mean_n_de_genes" = mean_mean_n_de,  #3.de genes
                   "max_mean_n_de_genes" = max_mean_n_de,
                   "min_mean_n_de_genes" = min_mean_n_de,
                   "mean_n_genes_lfc1" = mean_n_genes_lfc1,
                   "min_n_genes_lfc1" = min_n_genes_lfc1,
                   "max_n_genes_lfc1" = max_n_genes_lfc1,
                   "n_cells_total" = ncol(sce),              #4.general
                   "n_genes_total" = nrow(sce))


#Celltype-specificity? How celltype/cluster specific are batch effects? 
# Differences in size, distribution or abundance? Do we find correlations between lfcs,
# overlap in de genes, pathways? Interaction between ct and be?
celltype <- data.frame('mean_rel_abund_diff' = mean_rel_abund_diff, #1.abundance
                       'min_rel_abund_diff' = min_rel_abund_diff,
                       'max_rel_abund_diff' = max_rel_abund_diff,
                       "celltype_var_cms" = var_cms,                 #2.size/strength
                       "mean_de_overlap" = mean_de_overlap,
                       "min_de_overlap" = min_de_overlap,
                       "max_de_overlap" = max_de_overlap,
                       "mean_rel_cluster_spec"= mean_rel_spec,
                       "min_rel_cluster_spec"= min_rel_spec,
                       "max_rel_cluster_spec"= max_rel_spec,
                       "mean_lfc_cor" = mean_lfc_cor)


sim <- data.frame("mean_p_be" = mean_p_be,
                  "max_p_be" = max_p_be,
                  "min_p_be" = min_p_be,
                  "sd_p_be" = sd_p_be,
                  "mean_lfc_be" = mean_lfc_be,
                  "min_lfc_be" = min_lfc_be,
                  "max_lfc_be" = max_lfc_be,
                  "mean_p_ct"= mean_p_ct,
                  "min_p_ct"= min_p_ct,
                  "max_p_ct"= max_p_ct,
                  "sd_p_ct" = sd_p_ct)


summary <- cbind(size, celltype, sim) %>% set_rownames(dataset_name)

### -------------- save summary object ----------------------###
saveRDS(summary, file = outputfile)
```


```{r}
sessionInfo()
```

