---
title: "Visualize simulated batch effects"
author: "Almut Lütge"
date: "`r format(Sys.time(), '%d %B, %Y')`"
params:
  param_real:
  data:
  param:
  sim:
  real:
output:
  html_document:
    self_contained: no
    lib_dir: '../../docs/site_libs'
    code_folding: show
    theme: journal
    highlight: tango
    number_sections: no
    toc: yes
    toc_depth: 4
    toc_float: true
    collapsed: no
    smooth_scroll: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r dataset name, include=FALSE}
param <- readRDS(params$param)
dataset_name <- param[["dataset_name"]]

```

## `r paste(dataset_name)`

[Back to home](index.html)

```{r libraries, warning=FALSE}
suppressPackageStartupMessages({
  library(scater)
  library(CellMixS)
  library(purrr)
  library(data.table)
  library(here)
  library(tidyr)
  library(dplyr)
  library(stringr)
  library(ComplexHeatmap)
  library(edgeR)
  library(scran)
  library(cowplot)
  library(ggrepel)
  library(readr)
})
```


### Dataset and parameter
```{r data}
sce <- readRDS(params$data)
sce_real <- readRDS(params$real)
sim_list = list(strsplit(params$sim, ","))

print(sim_list[[1]][[1]])

param <- readRDS(params$param)
celltype <- param[["celltype"]]
batch <- param[["batch"]]
sample <- param[["Sample"]]
dataset_name <- param[["dataset_name"]]

param_real <- readRDS(params$param_real)
batch_real <- param_real[["batch"]]
cluster_real <- param_real[["celltype"]]
sample_real <- param_real[["sample"]]

table(colData(sce_real)[,sample_real])
#colors for plotting
tsne_cols <- c(
  "#DC050C", "#1965B0", "#FB8072","#7BAFDE", "#B17BA6",  
  "#FDB462", "#882E72", "#FF7F00", "#E78AC3", "#B2DF8A",
  "#33A02C", "#55A1B1", "#B2DF8A", "#8DD3C7", "#A6761D",
  "#7570B3", "#E6AB02", "#BEAED4", "#666666", "#999999")

mds_cols <- c(
  "#DC050C", "#1965B0", "#FB8072","#7BAFDE", "#B17BA6",  
  "#FDB462", "#882E72", "#FF7F00", "#E78AC3", "#B2DF8A",
  "#33A02C", "#55A1B1", "#B2DF8A", "#8DD3C7", "#A6761D",
  "#7570B3", "#E6AB02", "#BEAED4", "#666666", "#999999")
```


### Visualize simulations
```{r vis sim, warning=FALSE}

#Plot tsne
visUMAP <- function(simulation, title, ids){
  p <- visGroup(simulation, ids, dim_red = "UMAP") 
  p[["data"]]$cluster_id <- colData(simulation)[, "cluster_id"]
  p +  aes(shape = cluster_id) + 
      scale_shape_manual(values=1:nlevels(simulation$cluster_id)) +
      guides(colour = guide_legend(override.aes = list(size=3, alpha = 1))) +
      guides(shape = guide_legend(override.aes = list(size=3, alpha = 1), title = "celltype")) +
      geom_point(size=2, alpha = 1, aes_string(color="group_var")) +
      scale_color_manual(values = tsne_cols) +
      ggtitle(title)
}

visTSNE <- function(simulation, title, ids){
  p <- visGroup(simulation, ids)
  p[["data"]]$cluster_id <- colData(simulation)[, "cluster_id"]
  p + aes(shape = cluster_id) +
      scale_shape_manual(values=1:nlevels(simulation$cluster_id)) +
      guides(colour = guide_legend(override.aes = list(size=3, alpha = 1))) +
      guides(shape = guide_legend(override.aes = list(size=3, alpha = 1), title = "celltype")) +
      geom_point(size=2, alpha = 1, aes_string(color="group_var")) +
      scale_color_manual(values = tsne_cols) +
      ggtitle(title)
}

#Functions from muscat to generate pseudobulk sample and genrate a MDSplot
pbMDS <- function(x, title) {
    y <- as(assays(x), "list")
    y <- do.call("cbind", y)
    d <- suppressMessages(DGEList(y))
    d <- calcNormFactors(d)
    
    mds <- plotMDS.DGEList(d, plot = FALSE)
    ei <- metadata(x)$experiment_info
    m <- match(colnames(x), ei$sample_id)
    nk <- length(assays(x))
    df <- data.frame(
        MDS1 = mds$x, 
        MDS2 = mds$y, 
        cluster_id = rep(assayNames(x), each = ncol(x)),
        group_id = rep(ei$group_id[m], nk),
        sample_id = colnames(x))
    
    cols <- mds_cols
    if (nk > length(cols)) 
        cols <- colorRampPalette(cols)(nk)
    
    ggplot(df, aes_string(x = "MDS1", y = "MDS2", 
        col = "sample_id", shape = "cluster_id")) +
        scale_color_manual(values = cols) + 
        scale_shape_manual(values=1:nlevels(df$cluster_id)) +
        geom_point(size = 5, alpha = .8) + 
        guides(color = guide_legend(override.aes = list(alpha = 1))) +
        ggtitle(title) +
        theme_bw() + theme(aspect.ratio = 1,
            axis.text = element_text(color = "black"),
            panel.grid.minor = element_blank(),
            panel.grid.major = element_line(size = 0.2, color = "lightgrey"))
}

#helper functions from muscat
# Compute pseudo bulks
.pb <- function(x, cs, assay, fun) {
    fun <- switch(fun,
                  rowMedians = getFromNamespace(fun, "matrixStats"),
                  getFromNamespace(fun, "Matrix"))
    pb <- map_depth(cs, -1, function(i) {
        if (length(i) == 0) return(numeric(nrow(x)))
        fun(assays(x)[[assay]][, i, drop = FALSE])
    })
    map_depth(pb, -2, function(u) 
        data.frame(u, 
                   row.names = rownames(x),
                   check.names = FALSE))
}


#split cells
.split_cells <- function(x, 
                         by = c("cluster_id", "sample_id")) {
    if (is(x, "SingleCellExperiment"))
        x <- colData(x)
    cd <- data.frame(x[by], check.names = FALSE)
    cd <- data.table(cd, cell = rownames(cd)) %>% 
        split(by = by, sorted = TRUE, flatten = FALSE)
    map_depth(cd, length(by), "cell")
}

aggregateData <- function(x, assay,
    by = c("cluster_id", "sample_id"),
    fun = c("sum", "mean", "median"), 
    scale = FALSE) {
    
    if (missing("assay"))
        assay <- assayNames(x)[1]

    # validity checks for input arguments
    stopifnot(is.character(by), by %in% colnames(colData(x)))
    stopifnot(is.logical(scale), length(scale) == 1)
    
    # get aggregation function
    fun <- match.arg(fun)
    fun <- switch(fun,
        sum = "rowSums",
        mean = "rowMeans",
        median = "rowMedians")
    
    # split cells & compute pseudo-bulks
    cs <- .split_cells(x, by)
    pb <- .pb(x, cs, assay, fun)
    
    # scale
    if (scale) {
        if (assay == "counts" & fun == "rowSums") {
            pb_counts <- pb
        } else {
            pb_counts <- .pb(x, cs, assay, "rowSums")
        }
        pb <- map_depth(pb_counts, -2, function(u)
            u / colSums(u)[col(u)] * 1e6)
    }
    pb <- map_depth(pb, -2, as.matrix)
    
    # return SCE
    md <- metadata(x)
    md$agg_pars <- list(assay = assay, fun = fun)
    
    SingleCellExperiment(
        assays = pb,
        metadata = md)
}

visMDS <-function(simulation, title, ids){
  agg_sim <- aggregateData(simulation, by = c("cluster_id", ids),
    fun =  "mean", scale = TRUE)
  pbMDS(agg_sim, title)
}

```

### TSNE {.tabset}
```{r tsne, warning=FALSE, echo = FALSE}

template_tsne <- c(
    "#### {{nm}}\n",
    "```{r tsne{{nm}}, echo = FALSE}\n",
    "filename <- paste0('out/sim_char/', '{{name}}', '/sim_', '{{name}}', '_', '{{nm}}', '_sce.rds')",
    "sim_sce <- readRDS(file = filename)",
    "visTSNE(sim_sce, title = paste0('Simulate: ', '{{name}}', ' - ', '{{nm}}'), ids = batch)\n",
    "```\n",
    "\n"
  )

plots_tsne <- lapply(sim_list[[1]][[1]], 
  function(nm, name = dataset_name) knitr::knit_expand(text = template_tsne)
)
```

`r knitr::knit(text = unlist(plots_tsne))`


### UMAP {.tabset}
```{r umap, warning=FALSE}

template_umap <- c(
    "#### {{nm}}\n",
    "```{r umap{{nm}}, echo = FALSE}\n",
    "filename <- paste0('out/sim_char/', '{{name}}', '/sim_', '{{name}}', '_', '{{nm}}', '_sce.rds')",
    "sim_sce <- readRDS(file = filename)",
    "visUMAP(sim_sce, title = paste0('Simulate: ', '{{name}}', ' - ', '{{nm}}'), ids = batch)\n",
    "```\n",
    "\n"
  )

plots_umap <- lapply(sim_list[[1]][[1]], 
  function(nm, name = dataset_name) knitr::knit_expand(text = template_umap)
)
```

`r knitr::knit(text = unlist(plots_umap))`


### MDS plot {.tabset}
```{r mds, warning=FALSE}

template_mds <- c(
    "#### {{nm}}\n",
    "```{r mds{{nm}}, echo = FALSE}\n",
    "filename <- paste0('out/sim_char/', '{{name}}', '/sim_', '{{name}}', '_', '{{nm}}', '_sce.rds')",
    "sim_sce <- readRDS(file = filename)",
    "visMDS(sim_sce, title = paste0('Simulate: ', '{{name}}', ' - ', '{{nm}}'), ids = batch)\n",
    "```\n",
    "\n"
  )

plots_mds <- lapply(sim_list[[1]][[1]], 
  function(nm, name = dataset_name) knitr::knit_expand(text = template_mds)
)
```

`r knitr::knit(text = unlist(plots_mds))`

### Real data
```{r real}
# to make aggregate functions work
sce_real$batch_id <- as.factor(colData(sce_real)[, batch_real])
sce_real$sample_id <- as.factor(colData(sce_real)[, sample_real])
sce_real$cluster_id <- as.factor(colData(sce_real)[, cluster_real])
sce_real$group_id <- as.factor(rep("A", ncol(sce_real)))
metadata(sce_real)$experiment_info <- colData(sce_real)[, c("batch_id", "cluster_id", "group_id", "sample_id")]

visTSNE(sce_real, title = "Real dataset", ids = "batch_id")
visUMAP(sce_real, title = "Real dataset", ids = "batch_id")
visMDS(sce_real, title = "Real dataset", ids = "batch_id")
```


### Cellspecific Mixing score simulations {.tabset}  

```{r sim cms, echo = FALSE}
#get all cms scores
cms_all <- names(colData(sce))[grep("cms_sim", names(colData(sce)))]
cms_na <- colnames(colData(sce)[,cms_all])[colSums(is.na(colData(sce)[,cms_all])) > 0]
cms_all <- cms_all[!cms_all %in% cms_na]


template_cms <- c(
    "#### {{nm}}\n",
    "```{r cms{{nm}}, echo = FALSE}\n",
    "dim_nam <- gsub('cms_sim_','X', '{{nm}}')",
    "visMetric(sce, metric = '{{nm}}', dim_red = dim_nam)\n",
    "```\n",
    "\n"
  )

plots_cms <- lapply(cms_all, 
  function(nm) knitr::knit_expand(text = template_cms)
)

```

`r knitr::knit(text = unlist(plots_cms))`

